---
title: "Projects"
output:
  bookdown::html_document2:
    highlight: textmate
    toc: false
    toc_float:
      collapsed: true
      smooth_scroll: true
      print: false
    toc_depth: 4
    number_sections: false
    df_print: default
    code_folding: none
    self_contained: false
    keep_md: false
    encoding: 'UTF-8'
    css: "assets/lab.css"
    include:
      after_body: assets/footer-lab.html
---

```{r,child="assets/header-lab.Rmd"}
```

# &nbsp; {.tabset .tabset-fade}   

# Datasets  
Hands-on analysis of actual data is the best way to learn R programming. This page contains some data sets that you can use to explore what you have learned in this course. For each data set, a brief description as well as download instructions are provided. 

<div class="alert alert-info">
  <strong> Try to focus on using the tools from the course to explore the data, rather than worrying about producing a perfect report with a coherent analysis workflow.</strong>
</div>


On the last day you will present your Rmd file (or rather, the resulting html report) and share with the class what your data was about.

---

## Palmer penguins 🐧

- This is a data set containing a series of measurements for three species of penguins collected in the Palmer station in Antarctica.
- Data description: <https://vincentarelbundock.github.io/Rdatasets/doc/heplots/peng.html>

<details>
  <summary>Download instructions</summary>
```{r, warning=F, message=F}
penguins <- read.table("https://vincentarelbundock.github.io/Rdatasets/csv/heplots/peng.csv", header = T, sep = ",")
str(penguins)
```
</details>

---

## Drinking habits 🍷

- Data from a national survey on the drinking habits of american citizens in 2001 and 2002.
- Data description: <https://vincentarelbundock.github.io/Rdatasets/doc/stevedata/nesarc_drinkspd.html>

<details>
  <summary>Download instructions</summary>
```{r}
library(dplyr)
# this will download the csv file directly from the web
drinks <- read.table("https://vincentarelbundock.github.io/Rdatasets/csv/stevedata/nesarc_drinkspd.csv", header = T, sep = ",")
# the lines below will take a sample from the full data set
set.seed(seed = 2)
drinks <- sample_n(drinks, size = 3000, replace = F)
# and here we check the structure of the data
str(drinks)
```
</details>

---

## Car crashes 🚗

- Data from car accidents in the US between 1997-2002.
- Data description: <https://vincentarelbundock.github.io/Rdatasets/doc/DAAG/nassCDS.html>

<details>
  <summary>Download instructions</summary>
```{r}
library(dplyr)
# this will download the csv file directly from the web
crashes <- read.table("https://vincentarelbundock.github.io/Rdatasets/csv/DAAG/nassCDS.csv", header = T, sep = ",")
# the lines below will take a sample from the full data set
set.seed(seed = 2)
crashes <- sample_n(crashes, size = 3000, replace = F)
# and here we check the structure of the data
str(crashes)
```
</details>

---

## Gapminder health and wealth 📈

- This is a collection of country indicators from the Gapminder dataset for the years 2000-2016.
- Data description: <https://vincentarelbundock.github.io/Rdatasets/doc/dslabs/gapminder.html>

<details>
  <summary>Download instructions</summary>
```{r}
library(dplyr)
# this will download the csv file directly from the web
gapminder <- read.table("https://vincentarelbundock.github.io/Rdatasets/csv/dslabs/gapminder.csv", header = T, sep = ",")
# here we filter the data to remove anything before the year 2000
gapminder <- gapminder |> filter(year >= 2000)
# and here we check the structure of the data
str(gapminder)
```
</details>

---

## StackOverflow survey 🖥️

- This is a downsampled and modified version of one of StackOverflow's annual surveys where users respond to a series of questions related to careers in technology and coding.
- Data description: <https://vincentarelbundock.github.io/Rdatasets/doc/modeldata/stackoverflow.html>

<details>
  <summary>Download instructions</summary>
```{r}
library(dplyr)
# this will download the csv file directly from the web
stackoverflow <- read.table("https://vincentarelbundock.github.io/Rdatasets/csv/modeldata/stackoverflow.csv", header = T, sep = ",")
# the lines below will take a sample from the full data set
set.seed(2)
stackoverflow <- sample_n(stackoverflow, size = 3000)
# and here we check the structure of the data
str(stackoverflow)
```
</details>

---

## Doctor visits 🤒

- Data on the frequency of doctor visits in the past two weeks in Australia for the years 1977 and 1978.
- Data description: <https://vincentarelbundock.github.io/Rdatasets/doc/AER/DoctorVisits.html>

<details>
  <summary>Download instructions</summary>
```{r}
library(dplyr)
# this will download the csv file directly from the web
doctor <- read.table("https://vincentarelbundock.github.io/Rdatasets/csv/AER/DoctorVisits.csv", header = T, sep = ",")
# the lines below will take a sample from the full data set
set.seed(2)
doctor <- sample_n(doctor, size = 3000)
# and here we check the structure of the data
str(doctor)
```
</details>

---

## Video Game Sales 🎮

- This data set contains sales figures for video games titles released in 2001 and 2002.
- Data description: <https://mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=Video%20Game%20Sales> 
  - Click on "Preview Data" and "VG Data Dictionary" to see the description for each column.

<details>
  <summary>Download instructions</summary>
```{r, warning=F, message=F}
library(dplyr)
library(lubridate)
# this will download the file to your working directory
download.file(url = "https://maven-datasets.s3.amazonaws.com/Video+Game+Sales/Video+Game+Sales.zip", destfile = "video_game_sales.zip")
# this will unzip the file and read it into R
videogames <- read.table(unz(filename = "vgchartz-2024.csv", "video_game_sales.zip"), header = T, sep = ",", quote = "\"", fill = T)
# this will select rows corresponding to years 2001 and 2002
videogames <- filter(videogames, year(as_date(release_date)) %in% c(2001,2002))
# and here we check the structure of the data
str(videogames)
```
</details>

---

## LEGO Sets 🏗️

- This data set contains the description of all LEGO sets released from 2000 to 2009.
- Data description: <https://mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=lego>
  - Click on "Preview Data" and "VG Data Dictionary" to see the description for each column.

<details>
  <summary>Download instructions</summary>
```{r, warning=F, message=F}
library(dplyr)
# this will download the file to your working directory
download.file(url = "https://maven-datasets.s3.amazonaws.com/LEGO+Sets/LEGO+Sets.zip", destfile = "lego.csv.zip")
# this will unzip the file and read it into R
lego <- read.table(unz(filename = "lego_sets.csv", "lego.csv.zip"), header = T, sep = ",", quote = "\"", fill = T)
# this will select rows corresponding to years 2000-2009
lego <- filter(lego, year %in% seq(2000,2009,1))
# and here we check the structure of the data
str(lego)
```
</details>

---

## Shark attacks 🦈

- This data set contains information on shark attack records from all over the world.
- Data description: <https://mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=shark>
  - Click on "Preview Data" and "VG Data Dictionary" to see the description for each column.

<details>
  <summary>Download instructions</summary>
```{r, warning=F, message=F}
library(dplyr)
# this will download the file to your working directory
download.file(url = "https://maven-datasets.s3.amazonaws.com/Shark+Attacks/attacks.csv.zip", destfile = "attacks.csv.zip")
# this will unzip the file and read it into R
sharks <- read.table(unz(filename = "attacks.csv", "attacks.csv.zip"), header = T, sep = ",", quote = "\"", fill = T)
# the lines below will take a sample from the full data set
set.seed(seed = 2)
sharks <- sample_n(sharks, size = 3000, replace = F)
str(sharks)
```
</details>

***
# Sample project report  
<p><iframe style="border: none;" title="workshop" src="https://nbisweden.github.io/workshop-r/2410/Gapminder_project.html" width="100%" height="900px"></iframe></p>   
