---
title: "Loops in R"
subtitle: "R Programming Foundation for Life Scientists"
output:
  bookdown::html_document2:
    highlight: textmate
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
      print: false
    toc_depth: 4
    number_sections: true
    df_print: default
    code_folding: none
    self_contained: false
    keep_md: false
    encoding: 'UTF-8'
    css: "assets/lab.css"
    include:
      after_body: assets/footer-lab.html
editor_options: 
  chunk_output_type: console
---

```{r,child="assets/header-lab.Rmd"}
```

# Introduction

In programming languages loop structures, either with or without conditions, are used to repeat commands over multiple entities. `for` and `while` loops as well as `if`-`else` statements are also often used in R, but not as often as in many other programming languages. The reason for this is that many needs of the loops are addressed using **vectorization** or via `apply` functions.

This means that we can multiply all values in a vector in R by two by calling

```{r}
vec.a <- c(1, 2, 3, 4)
vec.a * 2
```

In many other and languages as well as in R, you can also create this with a loop instead

```{r}
for (i in vec.a) {
  vec.a[i] <- vec.a[i] * 2
}

vec.a
```

As you saw in the lecture, this is far less efficient and not by any means easier to type and we hence tend to avoid loops when possible.

After this exercise you should know:

- What are the most common loop structures in R
- Some common alternatives to using loops in R
- How one can convert a short script to a function.
- Use that new function in R.

# Exercises

1. Create a **100000 by 10** matrix with the numbers 1:1000000. Make a for-loop that calculates the sum for each row of the matrix. Verify that your results are consistent with what you obtain with the `apply()` function to calculate row sums as well as with the built-in `rowSums()` function. These functions were discussed in the lecture **Elements of the programming language - part 2**.

```{r,accordion=TRUE,results='markup'}
X <- matrix(1:1000000, nrow = 100000, ncol = 10)
for.sum <- vector()
# Note that this loop is much faster if you outside the loop create an empty vector of the right size.
# rwmeans <- vector('integer', 100000)
for (i in 1:nrow(X)) {
    for.sum[i] <- sum(X[i,])
}
head(for.sum)

app.sum <- apply(X, MARGIN = 1, sum)
head(app.sum)

rowSums.sum <- rowSums(X)
head(rowSums.sum)

identical(for.sum, app.sum)
identical(for.sum, rowSums.sum)
identical(for.sum, as.integer(rowSums.sum))
```

2. Another common loop structure that is used is the while loop, which functions much like a for loop, but will only run as long as a test condition is TRUE. Modify your for loop from exercise 1 and make it into a while loop.

```{r,accordion=TRUE}
x <- 1
while.sum <- vector("integer", 100000)
while (x < 100000) {
  while.sum[x] <- sum(X[x,])
  x <- x + 1
}
head(while.sum)
```

3. Create a data frame with two numeric and one character vector. Write a loop that loops over the columns and reports the sum of the column values if it is numeric and the total number of characters if it is a character vector.

```{r,accordion=TRUE}
vector1 <- 1:10
vector2 <- c("Odd", "Loop", letters[1:8])
vector3 <- rnorm(10, sd = 10)
dfr1 <- data.frame(vector1, vector2, vector3, stringsAsFactors = FALSE)
sum.vec <- vector()
for(i in 1:ncol(dfr1)) {
  if (is.numeric(dfr1[,i])) {
      sum.vec[i] <- sum(dfr1[,i])
  } else {
      sum.vec[i] <- sum(nchar(dfr1[,i]))
  }
}
sum.vec
```

4. In question 3 you generated a loop to go over a data frame. Try to convert this code to a function in R. The function should take a single data frame name as argument.

```{r,accordion=TRUE}
dfr.info <- function(dfr) {
sum.vec <- vector()
for (i in 1:ncol(dfr)) {
	if (is.numeric(dfr[,i])) {
		sum.vec[i] <- mean(dfr[,i])
	} else {
		sum.vec[i] <- sum(nchar(dfr[,i]))
	}
}
sum.vec
}
```

5.  Read up on the if-else function in R. If possible use the if-else function to answer question 3.

6.  In all loops that we tried out, we have created the variable where the output is saved outside the loop. Why is this?

7. <i class="fas fa-exclamation-circle"></i> **Advanced:** As explained in the lecture, an approach to calculate factorials were implemented using recursion (function calling itself). Here we instead will have a go at generating Fibonacci numbers. A fibonacci number is part of a series of number with the following properties:

The first two numbers in the Fibonacci sequence are either 1 and 1, or 0 and 1, depending on the chosen starting point of the sequence, and each subsequent number is the sum of the previous two. Hence:

`0, 1, 1, 2, 3, 5, 8, 13, 21, ...`

or

`1, 1, 2, 3, 5, 8, 13, 21, ...`

Try to generate such a series using a recursive approach.


