---
title: "Matrices, Lists and Dataframes"
subtitle: "R Programming Foundation for Life Scientists"
---

```{r,child="assets/header-lab.Rmd"}
```

# Introduction

A data set that have more than one dimension is conceptually hard to
store as a vector. For two-dimensional data set the solution to this
is to instead use matrices or data frames. As with vectors all values
in a matrix has to be of the same type (eg. you can not mix for
example characters and numerics in the same matrix). For data frames
this is not a requirement and different columns can have different
modes, but all columns in a data frame have the same number of
entries. In addition to these R also have objects named lists that can
store any type of data set and are not restricted by types or dimensions.

In this exercise you will learn how to:
- Create and work with matrices, data frames and lists
- Perform basic math operator on matrices
- Use functions to summarize information from data frames
- Extract subsets of data from macrices, data frames and lists
- Create S3 object from a list


# Matrices

The command to create a matrix in R is `matrix()`.
As input it takes a vector of values, the number of
rows and the number of columns.

```{r}
X <- matrix(1:12, nrow = 4, ncol = 3)
X
```

Note that if one only specify the number of rows or columns the it
will infer the size of the matrix automatically using the size of
vector and the option given. The default way of filling the matrix is
column-wise, so the first values from the vector ends up in column 1
of the matrix. If you instead wants to fill the matrix row by row you
can set the byrow flag to TRUE.

```{r}
X <- matrix(1:12, nrow = 4, ncol = 3, byrow = TRUE)
X
```

Subsetting a matrix is done the same way as for vectors, but you have
more than one dimension to work with. So you specify the rows and
column needed.

```{r}
X[1,2]
```

If one wants all values in a column or a row this can be specified by
leaving the other dimension empty, hence this code will print all
values in the second column.

```{r}
X[,2]
```

Note that if the retrieved part of a matrix can be represented as a
vector (eg one of the dimension have the length 1) R will convert it
to a vector otherwise it will still be a matrix.

## Exercise

Create a matrix containing 1:12 as shown similar to the matrix X above.

1.  How do find out the length and the mode of the matrix?

```{r,accordion=TRUE}
mode(X)
length(X)
```

2.  Extract all the values in the matrix that is larger than 6.

```{r,accordion=TRUE}
X[X>6]
```

3.  Swap the positions of column 1 and 3 in the matrix X

```{r,accordion=TRUE}
X[,c(3,2,1)]
```

4.  Add a vector with three zeros as a fifth row to the matrix

```{r,accordion=TRUE}
X.2 <- rbind(X, rep(0, 3))
X.2
```

5.  Replace all values the first two columns in your matrix with `NA`.

```{r,accordion=TRUE}
X[,1:2] <- NA
X
```

6.  Replace all values in the matrix with 0 and convert it to a vector

```{r,accordion=TRUE}
X[] <- 0
as.vector(X)
```

7.  In the the earlier exercises, you created a vector with the names of the type Geno\_a\_1, Geno\_a\_2, Geno\_a\_3, Geno\_b\_1, Geno\_b\_2&#x2026;, Geno\_s\_3 using vectors. In todays lecture a function named outer that generate matrices was mentioned. Try to generate the same vector as yesterday using this function instead. The outer function is very powerful, but can be hard to wrap you head around, so try to follow the logic, perhaps by creating a simple example to start with.

```{r,accordion=TRUE}
letnum <- outer(paste("Geno",letters[1:19], sep = "_"), 1:3, paste, sep = "_")
class(letnum)
sort(as.vector(letnum))
```

8.  Create two different 2 by 2 matrices named A and B. A should contain the values 1 - 4 and B the values 5-8. Try out the following commands and by looking at the results see if you can figure out what is going on. 

```
A. A * B
B. A / B
C. A %x% B
D. A + B
E. A - B
F. A == B
```

```{r,accordion=TRUE}
A <- matrix(1:4, ncol = 2, nrow = 2)
B <- matrix(5:8, ncol = 2, nrow = 2)
A
B

A * B
A / B
A %x% B
A + B
A - B
A == B
```

9. Generate a 10 by 10 matrix with random numbers. Add row and column names and calculate mean and median over rows and save these in a new matrix.

```{r,accordion=TRUE}
e <- rnorm(n = 100)
E <- matrix(e, nrow = 10, ncol = 10)
colnames(E) <- LETTERS[1:10]
rownames(E) <- colnames(E)
E.means <- rowMeans(E)
E.medians <- apply(E, MARGIN = 1, median)
E.mm <- rbind(E.means, E.medians)
E.mm
```

# Dataframes

Even though vectors are at the very base of R usage, data frames are central to R as the most common ways to import data into R (read.table) will create a data frame. Even though a data frame can itself contain another data frame, the by far, most common data frames consists of a set of equally long vectors. As data frames can contain several different data types the command `str()` is very useful to run on data frames

```{r}
vector1 <- 1:10
vector2 <- letters[1:10]
vector3 <- rnorm(10, sd = 10)
dfr <- data.frame(vector1, vector2, vector3)
str(dfr)
```

In the above example we can see that the dataframe **dfr** contains 10 observations for three variables that all have different modes, column 1 is an integer vector, column 2 a vector with factors! and column 3 a numeric vector. It is noteworthy that the second column is a factor even though we just gave it a character vector.

## Exercise

1. Figure out what is going on with the second column in **dfr** dataframe described above and modify the creation of the dataframe so that the second column is stored as a character vector rather than a factor. Hint: Check the help for `data.frame` to find an argument that turns off the factor conversion.

```{r,accordion=TRUE}
dfr <- data.frame(vector1, vector2, vector3, stringsAsFactors = FALSE)
str(dfr)
```

2. One can select columns from a data frame using either the name or the position. Use both methods to print the last two columns from the **dfr** data frame.

```{r,accordion=TRUE}
dfr[,2:3]
dfr[,c("vector2", "vector3")]
```

3. Print all letters in the **vector2** column of the data frame where the **vector3** column has a positive value.

```{r,accordion=TRUE}
dfr[dfr$vector3>0,2]
dfr$vector2[dfr$vector3>0]
```

4. Create a new vector combining the all columns of **dfr** separated by a underscore.

```{r,accordion=TRUE}
paste(dfr$vector1, dfr$vector2, dfr$vector3, sep = "_")
```

5. There is a dataframe of car information that comes with the base installation of R. Have a look at this data by typing `mtcars`. How many rows and columns does it have?

```{r,accordion=TRUE}
dim(mtcars)
ncol(mtcars)
nrow(mtcars)
```

6. Re-arrange the row names of this dataframe and save as a vector.

```{r,accordion=TRUE}
car.names <- sample(row.names(mtcars))
```

7. Create a dataframe containing the vector from the previous question and two vectors with random numbers named random1 and random2.

```{r,accordion=TRUE}
random1 <- rnorm(length(car.names))
random2 <- rnorm(length(car.names))
mtcars2 <- data.frame(car.names, random1, random2)
mtcars2
```

8. Now you have two dataframes that both contains information on a set of cars. A collaborator asks you to create a new dataframe with all this information combined. Create a merged dataframe ensuring that rows match correctly.

```{r,accordion=TRUE}
mt.merged <- merge(mtcars, mtcars2, by.x = "row.names", by.y = "car.names")
mt.merged
```

9. Calculate the mean value for the two columns that you added to the **mtcars** data frame.

```{r,accordion=TRUE}
colMeans(mtcars2[, c("random1", "random2")])
```

Try to modify so you get the mean by cylinder instead.

```{r,accordion=TRUE}
aggregate(mtcars2$random1, by=list(mtcars$cyl), FUN=mean)
```

# Lists

The last data structure that we will explore are lists, which is a very flexible structure. Lists can i R combine different data structures and they do not have to be of equal dimensions or have
other restrictions. The drawback with a flexible structure is that it requires a bit more work to interact with.

The syntax to create a list is similar to creation of the other data structures in R.

```{r}
l <- list(1, 2, 3)
```

As with the data frames the str() command is very useful for the sometimes fairly complex lists instances.

```{r}
str(l)
```

This example containing only numeric vector is not very exciting example given the flexibility a list structure offers so lets create a more complex example.

```{r}
vec1 <- letters
vec2 <- 1:4
mat1 <- matrix(1:100, nrow = 5)
df1 <- as.data.frame(cbind(10:1, 91:100))
u.2 <- list(vec1, vec2, mat1, df1, l)
```

As you can see a list can not only contain other data structures, but can also contain other lists.

Looking at the `str` command reveals much of the details of a list

```{r}
str(u.2)
```

With this more complex object, subsetting is slightly trickier than with more the more homogenous objects we have looked at so far.

To look at the first entry of a list one can use the same syntax as for the simpler structures, but note that this will give you a list of length 1 irrespective of the actual type of data structure found.

```{r}
u.2[1]
str(u.2[1])
```

If one instead wants to extract the list entry as the structure that is stored, one needs to "dig" deeper in the object.

```{r}
u.2[[1]]
str(u.2[[1]])
```

This means that the syntax to extract a specific value from a data structure stored in a list can be daunting. Below we extract the second column of a dataframe stored at position 4 in
the list **u.2**.  

```{r}
u.2[[4]][,2]
```

## Exercise

1. Create a list containing 1 character vector, a numeric vector, a character matrix.

```{r,accordion=TRUE}
list.2 <- list(vec1 = c("hi", "ho", "merry", "christmas"), vec2 = 4:19, mat1 = matrix(as.character(100:81), nrow = 4))
list.2
```

2. Create a dataframe and add this to the list.

```{r,accordion=TRUE}
dfr <- data.frame(letters, LETTERS, letters == LETTERS)
list.2[[4]] <- dfr
```

3. Remove the the second entry of your list.

```{r,accordion=TRUE}
list.2[-2]
```

4. Create a new list that contain 20 entries, with each entry holding a numeric vector.

```{r,accordion=TRUE}
vec1 <- rnorm(1000)
list.a <- split(vec1, 1:20)
```

5. How long is your list, and how long are each of the vectors that are part of the list?

```{r,accordion=TRUE}
length(list.a)
lapply(list.a, FUN = "length")
```

6. Figure out what the main differences are between the function `lapply()` and `sapply` are and use both of them with the function summary on your newly created list. What are the pros and cons of the two approaches to calculate the same summary statistics?

```{r,accordion=TRUE}
lapply(X = list.a, FUN = "summary")
sapply(X = list.a, FUN = "summary")
```

# Extras

1. Design a S3 class that should hold information on human proteins. The data needed for each protein is:

- The gene that encodes it
- The molecular weight of the protein
- The length of the protein sequence
- Information on who and when it was discovered
- Protein assay data

Create this hypothetical S3 object in R.

2. Among the test data sets that are part of base R, there is one called **iris**. It contains measurements on set of plants. You can access the data using by typing `iris` in R. Explore this data set and calculate some useful summary statistics, like SD, mean and median for the parts of the data where this makes sense. Calculate the same statistics for any grouping that you can find in the data.
