---
title: "Variables, Data types & Operators"
subtitle: "Elements of the R language"
author: "Marcin Kierczak & Nima Rafati"
keywords: bioinformatics, course, scilifelab, nbis, R
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/slide.css'
    lib_dir: libs
    include: NULL
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
---

exclude: true
count: false

```{r,echo=FALSE,child="assets/header-slide.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, width=60)
```

```{r,include=FALSE}
# load the packages you need
#library(dplyr)
#library(tidyr)
#library(stringr)
#library(ggplot2)
#library(mkteachr)
```

---
name: intro

# Introduction

In this lecture we will cover the basic elements of R language:  


- **variables and their types**
- **operators**
- vectors
- numbers as vectors
- strings as vectors
- matrices
- lists
- data frames
- objects
- repeating actions: iteration and recursion
- decision taking: control structures
- functions in general
- variable scope
- core functions

---
name: variables

# Variables

Imagine that you want to manipulate some numbers in your project simple operations can be done `2 + 2` but what if you want to do more complex operations or use the same number multiple times? 

```{r simple.addition,echo=TRUE,size='Tiny', eval=FALSE}
2 + 2
2 + 2 + 5
2 + 2 + 5 +10 
....
```
--

To use the same number multiple times you can use a **variable** to store the number and use the variable name instead of the number itself.  
We assign given value to a variable using `<-` operator (aka *assignment operator*).  

```{r assignment_operator,echo=TRUE,size='Tiny', eval=FALSE}
a <- 2 + 2
b <- 5 
c <- 5 + 10
sum_a_b <- a + b
sum_all <- a + b + c
```
--

# Is `<-` equivalent to `=`? Which one shall I use?  

- `val <- 3`, `val = 3` and `3 -> val` are three equivalent ways of assigning in R, But, it's best to use only `<-` to avoid possible confusion. The equal sign `=` should be used when setting function arguments ie; `f(a = 3)`.  

---
# Variables cted.

We are not constrained to numbers...

```{r variables2,echo=TRUE}
text1 <- 'a'
text2 <- 'qwerty'
text1  
text2
```


---
name: variables_naming

# Variables &mdash; naming conventions

- How to write variable names?
- What is legal/valid?
- What is a good style?

--

A syntactically valid name consists of letters, numbers and the dot or underline characters and starts with a letter or the dot not followed by a number.  

--

Names such as ".2way" are not valid, and neither are the so-called *reserved words*.  

---
name: variables_naming_reserved

# Reserved words

- `if, else, repeat, while, function, for, in, next, break, TRUE, FALSE, NULL, Inf, NaN, NA, NA_integer_, NA_real_, NA_complex_, NA_character_`  

- and you also **cannot** use: `c, q, t, C, D, I` and **SHOULD NOT** `T, F`

| Symbol | Function                   | Typical Use                        | 
| ------ | -------------------------- | ---------------------------------- | 
| `c`    | Combine values             | Create vectors                     | 
| `q`    | Quit session               | Exit R safely                      | 
| `t`    | Transpose                  | Matrix operations                  | 
| `C`    | Complex number constructor | Create complex values              | 
| `D`    | Derivative                 | Symbolic differentiation           | 
| `I`    | Inhibit interpretation     | Keep operators literal in formulas | 
| `T`   | Logical TRUE                | Logical operations                  |
| `F`   | Logical FALSE               | Logical operations                  |



---
name: variables_good_style

# Variables &mdash; good style

- So, how to name variables?

- make them informative, e.g. `genotypes` instead of `fsjht45jkhsdf4`,

--

- use consistent notation across your code &ndash; the same *naming convention*,

--

- camelNotation vs. dot.notation vs. dash_notation,

--

- do not `give_them_too_long_names`,

--

- in the dot notation avoid `my.variable.2`, use `my.variable2` instead,

--

- there are certain customary names:
    - `tmp` - for temporary variables;
    - `cnt` for counters;
    - `i,j,k` within loops,
    - `pwd` - for password...
--

**Why to care at all, I just want to analyze my data?**

- For your future self, coming back to the code after a while.  
- For your collaborators, who will read your code. 
- For the community, if you share your code publicly.  
---
name: data_types_1

# Variables have types

A *numeric* stores numbers of different *types*:

```{r variable.types1,echo=TRUE}
x <- 41.99 # assign 41.99 to x
typeof(x)
```

---
name: class_type_mode

# Classes, types, and modes

- `class` what type of object is it for R; **objects can be data (variables), functions, and models**.  
- `typeof()` what R thinks it is,

```{r class,echo=TRUE,eval=TRUE}
x <- 1:3
x
class(x)
typeof(x)

y <- 'hello'
y
class(y)
typeof(y)

lm <- lm(Sepal.Length ~ Sepal.Width, data=iris) # lm is a function for linear models. 
class(lm)
```

---
name: type_casting

# Type casting (Converting data types)  

By default, any *numeric* is stored as *double* !

```{r variable types2,echo=TRUE}
y <- 12 # now assign an integer value to y
class(y) # still numeric
typeof(y) # an integer, but still a double!
```

But we can explicitly **cast it** to integer:

```{r variable types3,echo=TRUE}
x <- as.integer(x) # type conversion, casting
typeof(x)
class(x)
is.integer(x)
```

> We need **casting** because sometimes a function requires data of certain type!

---
name: casting2

# More on type casting

Be careful when casting!

```{r careful.casting,echo=TRUE}
pi <- 3.1415926536 # assign approximation of pi to pi
pi
pi <- as.integer(pi) # not-so-careful casting
pi
pi <- as.double(pi) # trying to rescue the situation
pi
```
--

Casting is not rounding!

```{r casting.not.rounding,echo=TRUE}
as.integer(3.14)
as.integer(3.51)
```

---
name: ceiling_floor

# Ceiling, floor and a round corner

```{r casting.not.rounding2,echo=TRUE}
floor(3.51) # floor of 3.51
ceiling(3.51) # ceiling of 3.51
round(3.51, digits=1) # round to one decimal
```

---
name: coercion

# What happens if we cast a string to a number

```{r casting.string,echo=TRUE}
as.numeric('4.5678')
as.double('4.5678')
as.numeric('Nima is a cool teacher :) !')
```

---
name: some_special_values

# Special values

```{r special.values,echo=TRUE}
-1/0  # Minus infinity
1/0 # Infinity
```

and also:

```{r special.values2,echo=TRUE}
112345^67890  # Also infinity for R;  "^" means "to the power of"; 2^2 = 4
1/2e78996543  # Zero for R
Inf - Inf # Not a Number
```

---
name: logical_type

# Logical type

```{r logical.type,echo=TRUE}
a <- 7 > 2
b <- 2 >= 7
a
b
class(a)
typeof(a)
```
---
name: logical_cted

# Logical type cted.

```{r logical.type3,echo=TRUE}
x <- TRUE
x
x <- T # also valid
x
is.logical(x)
typeof(x)
```

- Observe that in R the logical type is also a numeric!

```{r logical.as.number,echo=TRUE}
x <- TRUE
y <- FALSE
x + y
2 * x
x * y
```

---
name: logical_joke

# A trap set up for you

Never ever use variable names as T or F. Why?

```{r TFtrap,echo=TRUE}
F <- T
T
F
```
Maybe applicable in politics, but not really in science...

---
name: character_type

# Character type

It is easy to work with characters and strings:

```{r char.type,echo=TRUE}
character <- 'c'
text <- 'This is my first sentence in R.'
text
character
class(character)
typeof(text) # also of 'character' type
```

---
name: character_type_2

# Character type

```{r char.type2,echo=TRUE, eval = FALSE}
number <- 3.14
number.text <- as.character(number) # cast to char
number.text
class(number.text)
as.numeric(number.text) # and the other way round
```
--

```{r char.type2_result,echo=FALSE, eval = TRUE}
number <- 3.14
number.text <- as.character(number) # cast to char
number.text
class(number.text)
as.numeric(number.text) # and the other way round
```
---
name: basic_string_ops

# Basic string operations

```{r basic.string,echo=TRUE}
  text1 <- "John had a yellow "
  text2 <- "submarine"
  result <- paste(text1, text2, ".", sep='')
  result
  sub("submarine", "cab", result)
  substr(result, start=1, stop=5)
```

---
name: basic_printing

# Basic printing

```{r basic.printing,echo=TRUE}
  txt <- "blue"
  (txt <- 'red')
  val <- 345.78
  sprintf("The weight of a %s ball is  %g g", txt, val)
```
---
# What did we learn?
<div style="font-size:70%"> 
 
| Concept            | Key Idea                                         | Example / Tip            |
| ------------------ | ------------------------------------------------ | ------------------------ |
| **Variables**      | Store data using names                           | `x <- 5`                 |
| **Assignment**     | Prefer `<-` for clarity                          | `a <- 2 + 2`             |
| **Naming rules**   | Must start with letter/dot; avoid reserved words | ❌ `if`, `c`, `T`, `F`    |
| **Good style**     | Use clear, consistent names                      | `gene_count`, `sampleID` |
| **Data types**     | Numeric, character, logical, etc.                | `typeof(x)` / `class(x)` |
| **Type casting**   | Change one type to another                       | `as.integer(3.14)`       |
| **Special values** | `Inf`, `-Inf`, `NaN`, `NA`                       | `1/0 → Inf`              |
| **Logical values** | TRUE/FALSE behave like 1/0                       | `TRUE + FALSE → 1`       |
| **Strings**        | Text manipulation                                | `paste("R", "rocks!")`   |
| **Operators**      | `+ − * / ^`  (power)                               | `2^3 → 8`                |
| **Objects**        | Everything in R is an object                     | data, functions, models  |

--
<div style="font-size:100%"> 
**Remember**
- R treats everything as an **object**.  
- Use `class()` to see what kind of object it is.  
- Use `typeof()` to see how it’s stored internally.  
- Code readability **>>** clever shortcuts.  
- Naming and clarity make collaboration easier.  

<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end_slide
class: end-slide, middle
count: false

# See you at the next lecture!

```{r, echo=FALSE,child="assets/footer-slide.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
#rmarkdown::render("presentation_demo.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("presentation_demo.html",output="presentation_demo.pdf")
```
