---
title: "Introduction To Programming in R (1)"
subtitle: "R Foundations for Life Scientists"
author: "Marcin Kierczak"
keywords: bioinformatics, course, scilifelab, nbis, R
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/slide.css'
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
---
exclude: true
count: false

```{r,echo=FALSE,child="assets/header-slide.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

```{r,echo=FALSE,message=FALSE,warning=FALSE}
# load the packages you need
#library(dplyr)
#library(tidyr)
#library(stringr)
#library(ggplot2)
```

---
name: programming_language

# Computer program

--

 * flow of data

--

 * from one function to another

--

 * each function does something to the data  

--

# How?

--
> A **programming language** is a formal computer language or constructed language designed to communicate instructions to a machine, particularly a computer. Programming languages can be used to create programs to control the behavior of a machine or to express algorithms.  
.small[[Wikipedia](https://en.wikipedia.org/wiki/Programming_language)]

--

# Three things to think about

 * what *types* of data can I process?

--

 * how do I *write* what I want?

--

 * when does it *mean* anything?


---

# Three components of a language

--

 * what *types* of data can I process &mdash; *type system*

--

  * int &mdash; 1 2 5 9
  * double &mdash; 1.23 -5.74
  * char &mdash; a b test 7 9

--

 * how do I *write* what I want &mdash; *syntax* defined by a language *grammar*

  `2 * 1 + 1` vs. `(+ (* 2 1) 1)`

--

 * when does it *mean* anything &mdash; *semantics*

--

  * *Colorful yellow train sleeps on a crazy wave.* &mdash; has no generally accepted meaning
  * *There is $500 on his empty bank acount.* &mdash; internal contradiction

---
name: topic2

# Where to start?


*Divide et impera* &mdash; divide and rule.

**Top-down approach:** define the big problem and split it into smaller ones. Assume you have solution to the small problems and continue &mdash; push the responsibility down.
Wishful thinking!

<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end_slide
class: end-slide, middle
count: false

# Thank you. Questions?

```{r,echo=FALSE,child="assets/footer-slide.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
rmarkdown::render("presentation_demo.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("presentation_demo.html",output="presentation_demo.pdf")
```
