---
title: "Brief introduction to statistics"
subtitle: "Statistics"
author: "Nima Rafati"
keywords: bioinformatics, course, scilifelab, nbis, R
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/slide.css'
    lib_dir: libs
    include: NULL
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
---

exclude: true
count: false

```{r,echo=FALSE,child="assets/header-slide.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, width=60)
```

```{r,include=FALSE}
# load the packages you need
#library(dplyr)
#library(tidyr)
#library(stringr)
#library(ggplot2)
#library(mkteachr)
```

---
name: intro2

# Introduction

**Why do we need statistics in our analysis?**  

- Make data understandable and insightful.  

- Evaluate patterns and trends.  

- Identify and quantify differences/similarities between groups. 


--


**Types of statistics:**  

- Descriptive statistics: To summarize and describe main features of a dataset (Mean, median,...).  

- Inferential statistics: To make prediction or inferences about a population using a sample of data (Hypothesis testing, regression analysis,...).  

- Predictive statistics: To make predictions about future outcomes based on collected data (Regression models, time series forecasting, machine learning,...). 

- ......  


---
name: intro2

In this lecture we will learn basic statistics concepts including:  

Descriptive statistics:  

- Central tendency: `mean`, `median`, `mode`.  
- Measures of spread: `range`, `interquartile range`, `variance`, `standard deviation`.  
- Correlation: `Pearson's correlation`, `Spearman's correlation`.  




---
name: Descriptive
# Types of Descriptive Statistics

Descriptive statistics helps to:

- Summarize and describe the data.  

- Visualize the data.   

- Identify patterns (trends) and outliers in the data.     

- Provide insights for downstream-analysis. 

---
name: Mean 
# Central Tendency: Mean
- Mean: The average value of data.  
$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$

```{r Mean, eval = T, echo = F, fig.width = 10, fig.height=4}
set.seed(123)
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2) + 0.1) 
data <- data.frame( var1 = rgamma(10000, shape = 2, scale = 2) * 12,
                    var2 = rnorm(10000, mean = 100, sd = 20))
hist(data$var1,breaks = 50, main = 'var1 distribution', xlab = 'var1', col = 'skyblue', freq = TRUE) 
var1_mean = mean(data$var1)
# Mean
abline(v = var1_mean, col = 'red', lwd = 2)
text(x = var1_mean + 10 , y = 700, labels = paste("Mean =", round(var1_mean, 2)), pos = 4, col = 'red', cex = 0.8)

hist(data$var2,breaks = 50, main = 'var2 distribution', xlab = 'var2', col = 'skyblue', freq = TRUE) 
var2_mean = mean(data$var2)
var2_median = median(data$var2)
# Mean
abline(v = var2_mean, col = 'red', lwd = 2)
text(x = var2_mean + 10 , y = 700, labels = paste("Mean =", round(var2_mean, 2)), pos = 4, col = 'red', cex = 0.8)
```

```{r mean, eval = T, echo = T}
mean(data$var1)
mean(data$var2)
```


---
name: Median 

# Central Tendency: Median

- Median: The middle value with the data is sorted.
```{r Median, eval = T, echo = F, fig.width = 10, fig.height=5}
par(mfrow=c(1,2))
hist(data$var1,breaks = 50, main = 'var1 distribution', xlab = 'var1', col = 'skyblue', freq = TRUE) 
var1_mean = mean(data$var1)
var1_median = median(data$var1)
# Mean
abline(v = var1_mean, col = 'red', lwd = 2)
text(x = var1_mean + 10 , y = 400, labels = paste("Mean =", round(var1_mean, 2)), pos = 4, col = 'red')
# Median
abline(v = var1_median, col = 'green', lwd = 2)
text(x = var1_median + 10 , y = 500, labels = paste("Median =", round(var1_median, 2)), pos = 4, col = 'green')

hist(data$var2,breaks = 50, main = 'var2 distribution', xlab = 'var2', col = 'skyblue', freq = TRUE) 
var2_mean = mean(data$var2)
var2_median = median(data$var2)
# Mean
abline(v = var2_mean, col = 'red', lwd = 2)
text(x = var2_mean + 30 , y = 400, labels = paste("Mean =", round(var2_mean, 2)), pos = 4, col = 'red')
# Median
abline(v = var2_median, col = 'green', lwd = 2)
text(x = var2_median + 30 , y = 600, labels = paste("Median =", round(var2_median, 2)), pos = 4, col = 'green')
```

```{r}
median(data$var1)
median(data$var2)
```

---
name: Mode
# Central Tendency: Mode

- Mode: The most frequently occurring value.    
```{r Mode-plot, eval = T, echo = F, fig.width = 10, fig.height=5}
par(mfrow=c(1,2))
hist(data$var1,breaks = 50, main = 'var1 distribution', xlab = 'var1', col = 'skyblue', freq = TRUE) 
var1_mean = mean(data$var1)
var1_median = median(data$var1)
# Mean
abline(v = var1_mean, col = 'red', lwd = 2)
text(x = var1_mean + 10 , y = 400, labels = paste("Mean =", round(var1_mean, 2)), pos = 4, col = 'red')
# Median
abline(v = var1_median, col = 'green', lwd = 2)
text(x = var1_median + 10 , y = 500, labels = paste("Median =", round(var1_median, 2)), pos = 4, col = 'green')
# Mode
density_data <- density(data$var1)
var1_mode <- density_data$x[which.max(density_data$y)]
abline(v = var1_mode, col = 'purple', lwd = 2)
text(x = var1_mode + 10 , y = 600, labels = paste("Mode =", round(var1_mode, 2)), pos = 4, col = 'purple')


hist(data$var2,breaks = 50, main = 'var2 distribution', xlab = 'var2', col = 'skyblue', freq = TRUE) 
var2_mean = mean(data$var2)
var2_median = median(data$var2)
# Mean
abline(v = var2_mean, col = 'red', lwd = 2)
text(x = var2_mean + 30 , y = 400, labels = paste("Mean =", round(var2_mean, 2)), pos = 4, col = 'red')
# Median
abline(v = var2_median, col = 'green', lwd = 2)
text(x = var2_median + 30 , y = 600, labels = paste("Median =", round(var2_median, 2)), pos = 4, col = 'green')
# Mode
density_data <- density(data$var2)
var2_mode <- density_data$x[which.max(density_data$y)]
abline(v = var2_mode, col = 'purple', lwd = 2)
text(x = var2_mode - 90 , y = 600, labels = paste("Mode =", round(var2_mode, 2)), pos = 4, col = 'purple')
```

```{r mode, echoo = T, eval = T}
density_data1 <- density(data$var1)
density_data1$x[which.max(density_data1$y)]
density_data2 <- density(data$var2)
density_data2$x[which.max(density_data2$y)]
```
---
name: Spread
# Measures of spread: Range and Interquartile Range. 
- Range: Difference between maximum `max(data$var2)`  and minimum `min(data$var2)`.  
- Interquartile Range: Data is represented in four equally sized groups (bins) known as **Quartile** and the distance between quartile is called **Interquartile Range** (IQR).    

```{r range, echo = F, eval = T}
# Sample data
set.seed(123)
data_quartile <- c(24, 30, 33, 45, 47, 58, 60, 66, 70)

# Calculate min, Q1, Q2 (median), Q3, max, IQR, and range
min_val <- min(data_quartile)
q1 <- quantile(data_quartile, 0.25)
median_val <- median(data_quartile)
q3 <- quantile(data_quartile, 0.75)
max_val <- max(data_quartile)
iqr_val <- IQR(data_quartile)
range_val <- max_val - min_val

# Plot the main line and quartiles
plot(c(1, 9), c(0, 1), type = "n", xlab = "", ylab = "", xaxt = "n", yaxt = "n", bty = "n")

# Main line (the range of the data)
segments(1, 0.5, 9, 0.5, lwd = 2)

# Draw vertical lines at min, Q1, median (Q2), Q3, max
segments(1, 0.45, 1, 0.55, lwd = 2) # Min
segments(3, 0.45, 3, 0.55, lwd = 2, col = "orange") # Q1
segments(5, 0.45, 5, 0.55, lwd = 2, col = "red") # Q2 (Median)
segments(7, 0.45, 7, 0.55, lwd = 2, col = "orange") # Q3
segments(9, 0.45, 9, 0.55, lwd = 2) # Max

# Add the values on top
text(1, 0.6, min_val, cex = 1)
text(3, 0.6, q1, cex = 1)
text(5, 0.6, median_val, cex = 1, col = "red")
text(7, 0.6, q3, cex = 1)
text(9, 0.6, max_val, cex = 1)

# Add labels for Min, Q1, Q2, Q3, Max
text(1, 0.4, "Min", cex = 1, col = "blue")
text(3, 0.4, "Q1", cex = 1, col = "blue")
text(5, 0.4, "Q2", cex = 1, col = "blue")
text(7, 0.4, "Q3", cex = 1, col = "blue")
text(9, 0.4, "Max", cex = 1, col = "blue")

# Add the IQR and Range arrows and labels
arrows(3, 0.3, 7, 0.3, length = 0.1)
text(5, 0.25, paste("IQR = Q3 - Q1 =", round(iqr_val, 2)), cex = 1)

arrows(1, 0.2, 9, 0.2, length = 0.1)
text(5, 0.15, paste("Range = Max - Min =", range_val), cex = 1)

```

---
name: Variance
# Measures of spread: Variance 

- Variance: How far the data points are spread out from the mean. Unit is the square of the data's unit (e.g. $cm^2$ ). 

$$
\sigma^2 = \frac{1}{n} \sum_{i=1}^{n} (x_i - \bar{x})^2
$$
```{r var, echo=TRUE, eval=TRUE}
var(data$var2)
```
---
name: Stdev
# Measures of spread: Standard deviation

- Standard deviation (sd): is the square root of the variance and provides a more intuitive measure of spread. Despite of variance, sd has the same unit as the data (e.g. cm). 

$$
\sigma =\sqrt{\sigma^2}
$$

```{r sd-plot,echo = F, eval = T}
var2_sd <- sd(data$var2)
hist(data$var2,breaks = 50, main = '', xlab = '', col = 'skyblue', freq = TRUE, ylim = c(0,1200))
abline(v = var2_mean, col = 'red', lwd = 2)
rect(var2_mean - var2_sd, 0, var2_mean + var2_sd, 1100, col = rgb(0.9, 0.9, 0.9, 0.5), border = NA)
rect(var2_mean - 2*var2_sd, 0, var2_mean - var2_sd, 1100, col = rgb(0.7, 0.7, 0.7, 0.5), border = NA)
rect(var2_mean + 2*var2_sd, 0, var2_mean + var2_sd, 1100, col = rgb(0.7, 0.7, 0.7, 0.5), border = NA)

rect(var2_mean - 3*var2_sd, 0, var2_mean - 2*var2_sd, 1100, col = rgb(0.5, 0.5, 0.5, 0.5), border = NA)
rect(var2_mean + 3*var2_sd, 0, var2_mean + 2*var2_sd, 1100, col = rgb(0.5, 0.5, 0.5, 0.5), border = NA)

text(x = var2_mean - 1 , y = 1200, labels = expression(bar(x)), pos = 4, col = 'red', cex = 0.8)
text(x = var2_mean + 5, y = 1100, labels = expression(bar(x) + sd), pos = 4, col = 'black', cex = 0.8)
text(x = var2_mean - var2_sd , y = 1100, labels = expression(bar(x) - sd), pos = 4, col = 'black', cex = 0.8)

text(x = var2_mean + 2*var2_sd - 15 , y = 1100, labels = expression(bar(x) + 2*sd), pos = 4, col = 'black', cex = 0.7)
text(x = var2_mean - 2*var2_sd , y = 1100, labels = expression(bar(x) - 2*sd), pos = 4, col = 'black', cex = 0.7)

text(x = var2_mean + 3*var2_sd - 15 , y = 1100, labels = expression(bar(x) + 3*sd), pos = 4, col = 'black', cex = 0.7)
text(x = var2_mean - 3*var2_sd , y = 1100, labels = expression(bar(x) - 3*sd), pos = 4, col = 'black', cex = 0.7)

rect(xleft = var2_mean - var2_sd,
     xright = var2_mean + var2_sd,
     ytop = 890,
     ybottom = 895) 
text(x = var2_mean + 30, y = 910, col = 'black', labels = '68.27%')
rect(xleft = var2_mean - 2*var2_sd,
     xright = var2_mean + 2*var2_sd,
     ytop = 590,
     ybottom = 595)
text(x = var2_mean + 40, y = 610, col = 'black', labels = '95.45%')
rect(xleft = var2_mean - 3*var2_sd,
     xright = var2_mean + 3*var2_sd,
     ytop = 190,
     ybottom = 195)
text(x = var2_mean + 50, y = 210, col = 'black', labels = '99.73%')

```
---
name: correlation
# Correlation

- Measuring the strength and direction of the **linear**  relationship between two variables.  

  - Positive Correlation: As one variable increases, the other also increases.
  
  - Negative Correlation: As one variable increases, the other decreases.
  
  - No Correlation: No directional relationship between the variables.
  
---
name: Pearson
# Types of correlation 
- Pearson's correlation coefficient: Correlation of two **continuous** variables.
- Assumptions:  
   - Linear relationship. 
   - Normally distributed variables.  

```{r pearson,echo = F, eval = T, fig.width=10, fig.height=5}
set.seed(123)

# Generate data for perfect positive correlation
x_pos <- seq(1, 100, length.out = 100)
y_pos <- x_pos + rnorm(100, mean = 0, sd = 1) # adding a tiny bit of noise for realism

# Generate data for perfect negative correlation
x_neg <- seq(1, 100, length.out = 100)
y_neg <- -x_neg + rnorm(100, mean = 0, sd = 1)

# Generate data for no correlation
x_none <- seq(1, 100, length.out = 100)
y_none <- rnorm(100, mean = 50, sd = 20)

# Combine all datasets into a data frame
data <- data.frame(
  x_pos = x_pos,
  y_pos = y_pos,
  x_neg = x_neg,
  y_neg = y_neg,
  x_none = x_none,
  y_none = y_none
)

# Plot the data to visualize the correlations
par(mfrow = c(1, 3), mar = c(5, 4, 4, 5) + 0.1)

# Positive correlation
plot(data$x_pos, data$y_pos, main = paste0("Positive  (r =", round(cor(data$x_pos, data$y_pos), digits = 4), ")"), xlab = "X", ylab = "Y", col = "blue", pch = 19)
abline(lm(data$y_pos ~ data$x_pos), col = "red", lwd = 2)

# Negative correlation
plot(data$x_neg, data$y_neg, main = paste0("Negative  (r =", round(cor(data$x_neg, data$y_neg), digits = 4), ")"), xlab = "X", ylab = "Y", col = "blue", pch = 19)
abline(lm(data$y_neg ~ data$x_neg), col = "red", lwd = 2)

# No correlation
plot(data$x_none, data$y_none, main = paste0("No Correlation (r =", round(cor(data$x_none, data$y_none), digits = 2), ")"), xlab = "X", ylab = "Y", col = "blue", pch = 19)
abline(lm(data$y_none ~ data$x_none), col = "red", lwd = 2)

```

$$
r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2}}
$$


---
name: Spearman
# Types of correlation 
- Spearman's rank correlation coefficient: Measures the monotonic relationship between two **ranked** variables.  
- Assumptions:  
  - It is a non-parametric approach and does not require the data to be linearly correlated.  
  - The data is not normally distributed.  
  - For both conrinuous and ordinal (categorical) variables.  
```{r spearman,echo = F, eval = T, fig.width=8, fig.height=4}
# Create the ordinal dataset
data_ordinal <- data.frame(
  Satisfaction = c(5, 4, 3, 2, 1, 4, 5, 2, 3, 1),
  Performance = c(9, 8, 7, 3, 2, 6, 10, 1, 5, 4)
)

# Calculate Spearman's rank correlation
spearman_corr <- cor(data_ordinal$Satisfaction, data_ordinal$Performance, method = "spearman")

# Plot to visualize the relationship
plot(data_ordinal$Satisfaction, data_ordinal$Performance,
     xlab = "Satisfaction (Ordinal)",
     ylab = "Performance (Rank)",
     main = paste("Spearman's Correlation =", round(spearman_corr, 2)),
     pch = 19, col = "blue")

# Add a line to show the trend
abline(lm(data_ordinal$Performance ~ data_ordinal$Satisfaction), col = "red", lwd = 2)

```

$$
\rho = 1 - \frac{6 \sum d_i^2}{n(n^2 - 1)}
$$
---
name: recap

<div style="font-size:60%"> 
# 🧮 Basic Statistics in R — Summary

| Concept | What it measures | Key R function(s) | Notes / Assumptions |
|----------|------------------|-------------------|----------------------|
| **Mean** | Average value (center) | `mean(x)` | Sensitive to outliers |
| **Median** | Middle value (robust center) | `median(x)` | Robust to skew/outliers |
| **Mode** | Most frequent value | `names(sort(table(x), TRUE))[1]` | Useful for discrete or multimodal data |
| **Range** | Full spread (max–min) | `range(x)` or `diff(range(x))` | Very sensitive to extremes |
| **IQR** | Middle 50% spread | `IQR(x)` | Robust to outliers |
| **Variance** | Average squared deviation from mean | `var(x)` | Units are squared; measures spread |
| **Standard Deviation (SD)** | Typical deviation from mean | `sd(x)` | Same unit as data; intuitive measure of spread |
| **Pearson’s r** | Linear relationship (−1 to 1) | `cor(x, y, method="pearson")` | Continuous, linear, normal assumption |
| **Spearman’s ρ** | Monotonic relationship (ranks) | `cor(x, y, method="spearman")` | Non-parametric; works for ranked or non-normal data |

<div style="font-size:100%"> 

# More on statistics?
- We discussed about very basic descriptive statistical measures.  
- You can read more [here](https://nbisweden.github.io/workshop-mlbiostatistics/session-descriptive/docs/index.html).    
<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end_slide
class: end-slide, middle
count: false

# See you at the next lecture!  
```{r, echo=FALSE,child="assets/footer-slide.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
#rmarkdown::render("presentation_demo.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("presentation_demo.html",output="presentation_demo.pdf")
```
