---
title: "Variables, data types, operators"
subtitle: "R Foundations for Life Scientists"
author: "`r paste0('<b>Marcin Kierczak</b> | ',format(Sys.time(), '%d-%b-%Y'))`"
institute: NBIS, SciLifeLab
keywords: bioinformatics, course, scilifelab, nbis, R
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/slide.css'
    lib_dir: libs
    include: NULL
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
---

exclude: true
count: false

```{r,echo=FALSE,child="assets/header-slide.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, width=60)
```

```{r,include=FALSE}
# load the packages you need
#library(dplyr)
#library(tidyr)
#library(stringr)
#library(ggplot2)
#library(mkteachr)
```

---
name: intro

## Introduction

Today, we will talk about various elements of a programming language and see how they are realized in R.

## Contents 

- **variables and their types**
- **operators**
- vectors
- numbers as vectors
- strings as vectors
- matrices
- lists
- data frames
- objects
- repeating actions: iteration and recursion
- decision taking: control structures
- functions in general
- variable scope
- core functions

---
name: variables

## Variables

> Creating a variable is simply assigning a name to some structure that stores data...

```{r variables,echo=TRUE,size='Tiny'}
7 + 9
a <- 7
a
b <- 9
b
c <- a + b
c
```

---
## Variables cted.
We are not constrained to numbers...
```{r variables2,echo=TRUE}
text1 <- 'a'
text2 <- 'qwerty'
text1  
text2
```

### Is `<-` equivalent to `=`? Which one shall I use?
> `val <- 3`, `val = 3` and `3 -> val` are in R three equivalent ways of assigning, BUT you should use only `<-` to avoid possible confusion. The equal sign `=` you can and should use 
when setting function arguments `f(a = 3)`.

---
name: variables_naming

## Variables &mdash; naming conventions

> How to write variable names? 

- What is legal/valid?
- What is a good style?

--

A syntactically valid name consists of letters, numbers and the dot or underline characters and starts with a letter or the dot not followed by a number.  

--

Names such as ".2way" are not valid, and neither are the so-called *reserved words*.  

--

### Reserved words
  
- `if, else, repeat, while, function, for, in, next, break, TRUE, FALSE, NULL, Inf, NaN, NA, NA_integer_, NA_real_, NA_complex_, NA_character_`  

- and you also **cannot** use: `c, q, t, C, D, I`

- and you **should not** use: `T, F`

---
name: variables_good_style

## Variables &mdash; good style

> So, how to name variables?

- make them informative, e.g. `genotypes` instead of `fsjht45jkhsdf4`,

--

- use consistent notation across your code &ndash; the same *naming convention*,

--

- camelNotation vs. dot.notation vs. dash_notation,

--

- do not `give_them_too_long_names`, 

--

- in the dot notation avoid `my.variable.2`, use `my.variable2` instead,

--

- there are certain customary names: 
    - `tmp` - for temporary variables; 
    - `cnt` for counters; 
    - `i,j,k` within loops, 
    - `pwd` - for password...

---
name: data_types_1

## Variables have types

A *numeric* stores numbers of different *types*:

```{r variable.types1,echo=TRUE}
x <- 41.99 # assign 41.99 to x
typeof(x)
```

---
name: class_type_mode

## Classes, types, and modes

- `class` what type of object is it for R,
- `typeof()` what R thinks it is,
- `mode()` how S language would see it (backward compatibility), 
- `storage.mode()` how is it stored in the memory; useful when talking to C or Java,

```{r class,echo=TRUE,eval=TRUE}
x <- 1:3
class(x)
typeof(x)
mode(x)
storage.mode(x)
```

---
name: type_casting

## Type casting

By default, any *numeric* is stored as *double* !

```{r variable types2,echo=TRUE}
y=12 # now assign an integer value to y
class(y) # still numeric
typeof(y) # an integer, but still a double!
```

But we can explicitly **cast it** to integer:

```{r variable types3,echo=TRUE}
x <- as.integer(x) # type conversion, casting
typeof(x)
class(x)
is.integer(x)
```

> We need **casting** because sometimes a function requires data of some type!

---
name: casting2

## More on type casting

Be careful when casting!

```{r careful.casting,echo=TRUE}
pi <- 3.1415926536 # assign approximation of pi to pi
pi
pi <- as.integer(pi) # not-so-careful casting
pi
pi <- as.double(pi) # trying to rescue the situation
pi
```

Casting is not rounding!

```{r casting.not.rounding,echo=TRUE}
as.integer(3.14)
as.integer(3.51)
```

---
name: ceiling_floor

## Ceiling, floor and a round corner

```{r casting.not.rounding2,echo=TRUE}
floor(3.51) # floor of 3.51
ceiling(3.51) # ceiling of 3.51
round(3.51, digits=1) # round to one decimal
```

---
name: coercion

## What happens if we cast a string to a number

```{r casting.string,echo=TRUE}
as.numeric('4.5678')
as.double('4.5678')
as.numeric('R course is cool!')
```

---
name: some_special_values

## Special values

```{r special.values,echo=TRUE}
-1/0  # Minus infinity
1/0 # Infinity
```

and also:

```{r special.values2,echo=TRUE}
112345^67890  # Also infinity for R
1/2e78996543  # Zero for R
Inf - Inf # Not a Number
```

---
name: cplx_numbers

## Complex number type

Core R supports complex numbers.

```{r cplx,echo=TRUE}
z = 7 + 4i # create a complex number
z
class(z)
typeof(z)
is.complex(z)
```

```{r cplx2,echo=TRUE} 
sqrt(-1) # not treated as cplx number
sqrt(-1 + 0i) # now a proper cplx number
sqrt(as.complex(-1)) # an alternative way
```

---
name: logical_type

## Logical type

```{r logical.type,echo=TRUE}
a <- 7 > 2
b <- 2 >= 7
a
b
class(a)
typeof(a)
```

R has three logical values: TRUE, FALSE and NA.

```{r logical.type2,echo=TRUE}
x <- c(TRUE, FALSE, NA)
names(x) <- as.character(x)
and_truth_table <- outer(x, x, "&") # AND table
```
```{r logical.DT, echo=FALSE}
kableExtra::kable(and_truth_table)
```

---
name: logical_cted

## Logical type cted.

```{r logical.type3,echo=TRUE}
x <- TRUE
x
x <- T # also valid
x
is.logical(x)
typeof(x)
```

> Observe that in R the logical type is also a numeric!

```{r logical.as.number,echo=TRUE}
x <- TRUE
y <- FALSE
x + y
2 * x
x * y
```

---
name: logical_joke

## A trap set up for you

Never ever use variable names as T or F. Why?

```{r TFtrap,echo=TRUE} 
F <- T
T
F
```
Maybe applicable in politics, but not really in science...

---
name: character_type

## Character type

It is easy to work with characters and strings:

```{r char.type,echo=TRUE}
character <- 'c'
text <- 'This is my first sentence in R.'
text
character
class(character)
typeof(text) # also of 'character' type
```

---
name: character_type_2

## Character type

```{r char.type2,echo=TRUE}
number <- 3.14
number.text <- as.character(number) # cast to char
number.text
class(number.text)
as.numeric(number.text) # and the other way round
```

---
name: basic_string_ops

## Basic string operations

```{r basic.string,echo=TRUE}
  text1 <- "John had a yellow "
  text2 <- "submarine"
  result <- paste(text1, text2, ".", sep='')
  result
  sub("submarine", "cab", result)
  substr(result, start=1, stop=5)
```

---
name: basic_printing

## Basic printing

```{r basic.printing,echo=TRUE}
  txt <- "blue"
  (txt <- 'red')
  val <- 345.78
  sprintf("The weight of a %s ball is  %g g", txt, val)
```

<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end_slide
class: end-slide, middle
count: false

# See you at the next lecture!

```{r, echo=FALSE,child="assets/footer-slide.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
#rmarkdown::render("presentation_demo.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("presentation_demo.html",output="presentation_demo.pdf")
```

