---
title: "Reading (and writing) data in R"
subtitle: "R Foundations for Data Analysis"
author: "Marcin Kierczak"
keywords: "bioinformatics, course, scilifelab, nbis, R"
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/slide.css'
    lib_dir: libs
    include: NULL
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
---

exclude: true
count: false

```{r,echo=FALSE,child="assets/header-slide.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, width=60)
```

```{r,echo=FALSE,message=FALSE,warning=FALSE}
# load the packages you need
#library(dplyr)
#library(tidyr)
#library(stringr)
#library(ggplot2)
#library(mkteachr)
```

---
name: reading_data

# Reading data

* Can be one of the most consuming and cumbersome aspects of data analysis.

* R provides ways to read and write data stored on different media (e.g.: file, database, url) and in different formats.

* Package `foreign` contains a number of functions to import less common data formats.

---
name: reading_tables

# Reading tables

We can use the `read.table()` function. It is a nice way to read your data into a data frame.  

The function is declared in the following way:  

```{r, echo=T, eval=F}
read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = FALSE,
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)

# or just
read.table(file)
```

---
name: read_table_params

# `read.table()` parameters

You can read all about the `read.table()` function using `?read.table`

The most important arguments are:

* **file** &ndash; the path to the file that contains data, e.g. `/path/to/my/file.csv`
* **header** &ndash; a logical indicating whether the first line of the file contains variable names,
* **sep** &ndash; a character determining variable delimiter, e.g. `","` for csv files,
* **quote** &ndash; a character telling R which character surrounds strings,
* **dec** &ndash; character determining the decimal separator,
* **row/col.names** &ndash; vectors containing row and column names,
* **na.strings** &ndash; a character used for missing data,
* **nrows** &ndash; how many rows should be read,
* **skip** &ndash; how many rows to skip,
* **as.is** &ndash; a vector of logicals or numbers indicating which columns shall not be converted to factors,
* **fill** &ndash; add NA to the end of shorter rows,
* **stringsAsFactors** &ndash; a logical. Rather self explanatory.

---
name: read_table_sibs

# `read.table` and its siblings  

`read.table` has some variants with presets to specific file types:

* `read.csv()` and `read.csv2()` with comma and semicolon as default `sep`, and dot and comma as `dec` respectively,
* `read.delim()` and `read.delim2()` for reading tab-delimited files.


---
name: read_table_example

# `read.table` &mdash; example use

```{r read.table, echo=T}
tab <- read.table(file = 'data/slide_loading_data/2014-04-07_phenos2.csv',
                  sep = ' ',
                  header = T)
tab[1:5, 1:3]
class(tab$reg_no)
```

---
name: handling_errors

# What if you encounter errors?

* R documentation `?` and `??`
* Google &ndash; just type R and copy the error you got without your variable names,
* Open the file using a text editor and see if you can spot anything unusual &ndash; 
  * e.g. does the header line have the same number of columns as the first line?

--

# Useful terminal commands for debugging (Linux/OsX)

* `cat phenos.txt | awk -F';' '{print NF}'` prints the number of words in each row. `-F';'` says that semicolon is the delimiter,

* `head -n 5 phenos.txt` prints the 5 first lines of the file,

* `tail -n 5 phenos.txt` prints the 5 last lines of the file,

* `head -n 5 phenos.txt | tail -n 2` will print lines 4 and 5...

* `wc -l phenos.txt` will print the number of lines in the file

* `head -n 2 phenos.txt > test.txt` will write the first 2 lines to a new file

--

If it still does not give you a clue &mdash; just try to load first line of the file.  

--

If this still did not help, split the file in two equal-size parts. Check which part gives the error. Split this part into halves and check which 1/4 gives the error... It is faster than you think!

---
name: writing

# Writing with `write.table()`

`read.table()` has its counterpart, the `write.table()` function (as well ass its siblings, like `write.csv()`). You can read more about it in the documentation, let us show some examples:

```{r write.table, echo=T, eval=F}
vec <- rnorm(10)
write.table(vec, '') # write to screen
write.table(vec, file = 'vector.txt')
```

```{r, echo=F,eval=F}
# write to the system clipboard, handy!
write.table(vec, 'clipboard', col.names=F, row.names=F)

# or on OsX
clip <- pipe("pbcopy", "w")                       
write.table(vec, file=clip)                               
close(clip)

# To use in a spreadsheet
write.csv(vec, file = 'spreadsheet.csv')
```

---
name: write_big_data

# Writing big data

* HINT: `write.table()` is rather slow on big data &ndash; it checks types for every column and row and does separate formatting to each. If your data consists of only one type of data, convert it to a matrix using `as.matrix` before you write it!  

* You may want to use function `scan()` that reads files as vectors. The content does not have to be in the tabular form. You can also use scan to read data from keyboard: `typed.data <- scan()`

* If data are written as fixed-width fields, use the `read.fwf()` function.

* Also check out the `readLines()` function that enables you to read data from any stream.

---
name: read_xls_matlab

# Read data in xls/xlsx and Matlab

```{r xls, eval=F, echo=T}
library(readxl)
data <- read_xlsx('myfile.xlsx')
```

```{r matlab, eval=F, echo=T}
library(R.matlab)
data <- readMat("mydata.mat")
```

---
name: remote_data

# Working with remote data

```{r url.data, eval=F, cache=T, echo=T}
url <- 'https://en.wikipedia.org/wiki/List_of_countries_by_average_wage'
conn <- url(url, 'r')
raw.data <- readLines(conn)
raw.data[1:3]
```

But data is often tabularized...

```{r url.data.2, eval=F, cache=T, echo=T}
library(rvest)
html <- read_html(url)
tables <- html_nodes(html, 'table')
data <- html_table(tables[4])[[1]]
data[1:5, ]
```

---
name: databases

# Working with databases

It is also relatively easy to work with different databases. We will focus on MySQL and present only one example that uses the *RMySQL* package (check also *RODBC* and *RPostgreSQL*).
```{r db, echo=T, eval=F}
library(RMySQL)
db.conn <- dbConnect(MySQL(), user='me',
                     password='qwerty123',
                     dbname='genes',
                     host='127.0.0.237')
query <- dbSendQuery(db.conn, 'SELECT * FROM table7')
data <- fetch(query, n = - 1)
```

---
name: capabilities

# Capabilities

If you are getting some errors, e.g. trying to connect to a url, you may check whether your system (and R) support particular type of file or connection:
```{r capabil, echo=T, size='tiny'}
capabilities()
```

<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end_slide
class: end-slide, middle
count: false

# See you at the next lecture!

```{r, echo=FALSE,child="assets/footer-slide.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
#rmarkdown::render("presentation_demo.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("presentation_demo.html",output="presentation_demo.pdf")
```
